The symmetry of partner modelling

### 2016-02-04 mutual-modeling.tex
Updated section 1

From its very beginning, CSCL research has been following
the suggestion by \citet{roschelle1995construction} that collaborative learning has
something to do with the process of constructing and maintaining a \emph{shared
understanding} of the task at hand. Building a shared/mutual understanding
refers to the upper class of collaborative learning situations, those in which
students should build upon each other's understanding to refine their own
understanding: what is expected to produce learning is not the mere fact that
two students build the same understanding but the cognitive effort they have to
engage to build this shared understanding~\citep{schwartz1995emergence}. This
effort can be observed by the frequency of rich interactions, \ie interactions
whose occurrence has been related to learning: (self-) explanations in cognitive
science~\citep{chi1989self, webb1991task}, conflict resolution in
socio-cognitive theories~\citep{doise1975social},
mutual regulation~\citep{blaye1995collaborative} in a Vygostkian perspective, etc. The
construction of a shared understanding has been investigated for several decades
in psycholinguistics, under the  notion of
\emph{grounding}~\citep{clark1986referring}. However, the relevance of grounding
mechanisms for explaining learning outcomes has been questioned in the learning
sciences. The monitoring and repair of misunderstanding explains for instance
referential failures in short dialogue episodes but does hardly predict
\emph{conceptual change} (\ie the acquisition, acceptation and integration of a
new belief into one's mental model) over longer
sessions~\citep{dillenbourg2006sharing}. The cumulative effect of grounding
episodes can probably be better understood from a socio-cultural perspective:

@@

Paradoxically, while Clark's theory is somewhat too linguistic from a conceptual
change viewpoint, it is criticized at the same time as being too cognitivist by
some psycholinguists, \ie as overestimating the amount of shared knowledge and
mutual representations actually necessary to conduct a dialogue. The fundamental
issue, as old as philosophy, is the degree of coupling between the different
levels of dialogue, mostly between the lexical/syntactical level and the deeper
semantic levels. \citet{pickering2006alignment} argue that the mutual
understanding starts mostly with a \emph{superficial alignment} at the level of
the linguistic representations, due to priming mechanisms, and that this local
alignment may -- in some cases -- lead to a \emph{global alignment} of the
semantic level (\emph{deep grounding}).  For these authors, the convergence in
dialogue, and even the repair of some misunderstandings, is explained by this
mimetic behavior more than by a monitoring of each other's knowledge:
\emph{``...interlocutors do not need to monitor and develop full common ground
as a regular, constant part of routine conversation, as it would be
unnecessary and far too costly. Establishment of full common ground is, we
argue, a specialized and non-automatic process that is used primarily in
times of difficulty (when radical misalignment becomes apparent).''} (p. 179).
This view is actually not incompatible with Clark's \emph{grounding
criterion}~\citep{clark1989contributing}: the degree of shared understanding that
peers need to reach depends upon the task they perform. For instance, a dialogue
between two surgeons might rely on superficial alignment if they talk about
their friends but has to guarantee accurate common grounds when talking about
which intervention will be conducted in which way on which patient. 
%Some CSCL
%scholars stressed the ``illusion of shared understanding'' \fixme{REF: Jacco?},
%\ie the potential decoupling between linguistic alignment and actual shared
%meanings.

This interesting cognitive science debate occurred mostly outside the field of
learning. In education, the question is to relate these mechanisms to learning
outcomes: \textbf{Is linguistic alignment sufficient to trigger conceptual
change?} \textbf{Does negotiation of meaning only occur when partners monitor
and diagnose each other's knowledge?} If the ratio between shallow alignment and
deep grounding depends upon the task, and if deep grounding is a condition for
learning, then the pedagogical challenge is indeed to design tasks that require
deep grounding. Most empirical studies on grounding and alignment are conducted
with tasks that, despite being qualified as ``ecologically valid'' by their
authors, are mere referencing tasks such as asking the way to the train station
or helping the peer to choose a picture among many. In this contribution, we
explore several richer tasks such as arguing on a sensitive issue or building a
concept map.

@@

\vspace{2em}
These preliminary remarks frame the main question that this article address:
in order to engage into shared understanding, which is a desirable learning
situation, do partners have to build a representation of each other's knowledge?
if so, what exactly is the nature of this mutual modelling? We adopt hereafter a
constructive approach, first introducing a simple yet convenient model of mutual
modelling, then analyzing five independent studies in terms of the interactions
between learning and mutual modelling.

>>>>

From its very beginning, CSCL research has been following
the suggestion by \citet{roschelle1995construction} that collaborative learning has
something to do with the process of constructing and maintaining a \emph{shared
understanding} of the task at hand. Building a shared/mutual understanding
refers to the upper class of collaborative learning situations, those in which
students should build upon each other's understanding to refine their own
understanding: what is expected to produce learning is not the mere fact that
two students build the same understanding but the cognitive effort they have to
engage to build this shared understanding~\citep{schwartz1995emergence}. This
effort can be observed by the frequency of rich interactions, \ie interactions
whose occurrence has been related to learning: (self-) explanations in cognitive
science~\citep{chi1989self, webb1991task}, conflict resolution in
socio-cognitive theories~\citep{doise1975social},
mutual regulation~\citep{blaye1995collaborative} in a Vygostkian perspective, etc. The
construction of a shared understanding has been investigated for several decades
in psycholinguistics, under the  notion of
\emph{grounding}~\citep{clark1986referring}. However, the relevance of grounding
mechanisms for explaining learning outcomes has been questioned in the learning
sciences. Grounding mechanisms  are appropriate for explains conversational events, such
referential failures in short dialogue episodes, but they do  hardly predict deeper phenomena such as
\emph{conceptual change} (\ie the acquisition, acceptation and integration of a
new belief into one's mental model) over longer
sessions~\citep{dillenbourg2006sharing}. The cumulative effect of grounding
episodes can probably be better understood from a socio-cultural perspective:

@@

Paradoxically, while Clark's theory is somewhat too linguistic from a learning  viewpoint, it is criticized at the same time as being too cognitivist by
some psycholinguists, \ie as overestimating the amount of shared knowledge and
mutual representations actually necessary to conduct a dialogue. The fundamental
issue, as old as philosophy, is the degree of coupling between the different
levels of dialogue, mostly between the lexical/syntactical level and the deeper
semantic levels. \citet{pickering2006alignment} argue that the mutual
understanding starts mostly with a \emph{superficial alignment} at the level of
the linguistic representations, due to priming mechanisms, and that this local
alignment may -- in some cases -- lead to a \emph{global alignment} of the
semantic level (\emph{deep grounding}).  For these authors, the convergence in
dialogue, and even the repair of some misunderstandings, is explained by this
mimetic behavior more than by a monitoring of each other's knowledge:
\emph{``...interlocutors do not need to monitor and develop full common ground
as a regular, constant part of routine conversation, as it would be
unnecessary and far too costly. Establishment of full common ground is, we
argue, a specialized and non-automatic process that is used primarily in
times of difficulty (when radical misalignment becomes apparent).''} (p. 179).
This view is actually not incompatible with Clark's \emph{grounding
criterion}~\citep{clark1989contributing}: the degree of shared understanding that
peers need to reach depends upon the task they perform. For instance, a dialogue
between two surgeons might rely on superficial alignment if they talk about
their friends but has to guarantee accurate common grounds when talking about
which intervention will be conducted in which way on which patient.  In this paper, we operationalized the grounding criterion, i.e. the necessity for accurate modelling, as the correlation between the accuracy of partner models and various measures of team performance.

This interesting cognitive science debate occurred mostly outside the field of
learning. In education, the question is to relate these mechanisms to learning
outcomes: \textbf{Is linguistic alignment sufficient to trigger conceptual
change?} \textbf{Does negotiation of meaning only occur when partners monitor
and diagnose each other's knowledge?} If the ratio between shallow alignment and
deep grounding depends upon the task, and if deep grounding is a condition for
learning, then the pedagogical challenge is indeed to design tasks that require
deep grounding. Most empirical studies on grounding and alignment are conducted
with simple referencing tasks such as asking the way to the train station
or helping the peer to choose a picture among many. In the studies we report here, we
explore several richer tasks such as arguing on a sensitive issue or building a
concept map.

@@

\vspace{2em}
In summary, the controversy around the cognitive depth of shared understanding pertains to psycholinguistic, which investigates 'natural conversations, but  in collaborative learning, the situation is different; tasks are precisely designed for requirint a deeper negotiation fo meaning.  The craft of CSCL is to design collaborative situations that create interdependence, avoid groupthink, and allows students to detect any 'illusion of shared understanding' (Cherubini et al, 2005). Our question is hence not  anymore"do peers build a shared understanding ?", but becomes "when peers have to build a shared understanding, how do they do it ?".  This question addresses the mechanisms of grounding, in which the basic sequence is: make a proposition, detect misunderstanding and, if any, repair them.  This paper focuses on the middle part, the detection of misunderstandings.  Detecting a misunderstanding means that the emitter of a message identifies a mismatch between his communicative intention and the way his message is understood by his or her partner. Detecting one peer's understanding is investigate hereafter under the general umbrella of partner modelling.

Cherubini, M., Van Der Pol, J., & Dillenbourg, P. (2005). Grounding is not shared understanding: Distinguishing grounding at an utterance and knowledge level. In CONTEXT'05, the Fifth International and Interdisciplinary Conference on Modeling and Using Context (pp. 11-23).

%These preliminary remarks frame the main question that this article address:
%in order to engage into shared understanding, which is a desirable learning
%situation, do partners have to build a representation of each other's knowledge?
%if so, what exactly is the nature of this mutual modelling? We adopt hereafter a
%constructive approach, first introducing a simple yet convenient model of mutual
%modelling, then analyzing five independent studies in terms of the interactions
%between learning and mutual modelling.
